Covariate analysis of UFL-Mayo-ISB RNASeq reprocessed counts (both cerebellum and temporal cortex)

### 2016-07-13 code/Rmd/Mayo_RNASeq_Reprocessed.Rmd
Included PMI predicitons based on the availability of PMI values

## Load required libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

@@

@@

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% c("11386_CER","1923_CER", "11311_CER", "11477_CER", "11408_TCX", "11294_TCX", "11396_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance:
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

@@

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project. Following samples were removed `r indRemoved`

@@

multiplot(plotlist = p, cols = 2)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")

@@

COVARIATES.tmp = COVARIATES
COVARIATES.tmp[,FactorCovariates] = lapply(COVARIATES.tmp[,FactorCovariates], as.numeric)

WGCNA::plotDendroAndColors(sampleTrees, 
                           labels2colors(cbind(tree,COVARIATES.tmp)),
                           groupLabels = c('Cluster',colnames(COVARIATES.tmp)))

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
postAdjustCovars = c('BrainRegion.Diagnosis')
randomCovariates = 'Donor_ID'

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`

@@

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
plotDendroAndColors(sampleTrees, 
                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp)), 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[clust == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[clust == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 4 are', paste(sampleTrees$labels[clust == 4], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```

@@

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
write.table(COVARIATES, file = 'MAYOcovariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MAYOcovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

@@

# Design Matrix
write.table(DM1$design, file='MAYOdesign.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MAYOdesign.tsv', 
                  name = 'Design Matrix',
                  parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ,
                      used = ALL_USED_IDs, 
                      activityName = activityName, 
                      executed = thisFile, 
                      activityDescription = activityDescription)

>>>>

## Load required libraries
library(CovariateAnalysis) ## install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

@@

library(glmnet)

@@

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% c("11386_CER","1923_CER", "11311_CER", "11477_CER", "11408_TCX", "11294_TCX", "11396_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: `r paste(c("11386_CER","1923_CER", "11311_CER", "11477_CER", "11408_TCX", "11294_TCX", "11396_TCX"), collapse = ',')`
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

@@

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project. Following samples were removed `r indRemoved`
### Predict missing PMI
```{r predict.pmi}
PMI = read.table('Mayo_PMI.tsv', header=T, fill=T) %>%
  dplyr::select(-Source) %>% 
  right_join(METADATA %>%
               dplyr::mutate(SubjectId_Synpase = paste(Donor_ID, BrainRegion, sep = '_')))
rownames(PMI) = PMI$SubjectId_Synpase

# Get genes that are predictors of PMI and use them as control probes
CONTROL.PROBES_ID = 'syn6145639'
ALL_USED_IDs = c(ALL_USED_IDs, CONTROL.PROBES_ID)
CONTROL.PROBES = fread(synGet(CONTROL.PROBES_ID)@filePath, data.table = F)

# Remove genes
EXPR <- getGeneFilteredGeneExprMatrix(COUNT, 
                                      MIN_GENE_CPM=0,
                                      MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0,
                                      verbose=T)

# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=F)

ind = sapply(rownames(VOOM.GENE_EXPRESSION$E), function(x){str_split(x, '\\.')[[1]][1]}) %in% CONTROL.PROBES$x 
EXPR = VOOM.GENE_EXPRESSION$E[ind,]
PMI = PMI[colnames(EXPR),]

expr = dlply(PMI, .(Donor_ID), function(mtd, expr){
  if (dim(mtd)[1] > 1)
    expr = rowMeans(expr[, mtd$SubjectId_Synpase]) %>% rownameToFirstColumn('GeneID')
  else
    expr = expr[, mtd$SubjectId_Synpase] %>% rownameToFirstColumn('GeneID')
  expr = expr %>% plyr::rename(c('DF' = unique(mtd$Donor_ID)))
  return(expr)
}, EXPR) %>%
  join_all
rownames(expr) = expr$GeneID

PMI = PMI[,colnames(PMI) %in% c('Donor_ID','PMI')] %>% unique
rownames(PMI) = PMI$Donor_ID

PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
expr = expr[,intersect(rownames(PMI), colnames(expr))]

# Split data into trainining and prediction set
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]

plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('Donor_ID') %>%
  plyr::rename(c('1' = 'PMI'))

PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)

METADATA = left_join(METADATA, PMI)
```

@@

# PMI
p[[3]] = ggplot(METADATA, aes(x = Diagnosis, y = as.numeric(PMI), fill = Tissue)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')

multiplot(plotlist = p, cols = 3)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")

@@

COVARIATES.tmp = COVARIATES
COVARIATES.tmp[,FactorCovariates] = lapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

WGCNA::plotDendroAndColors(sampleTrees, 
                           col.mat,
                           groupLabels = c('Cluster',colnames(COVARIATES.tmp)))

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrainRegion.Diagnosis')
randomCovariates = 'Donor_ID'

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Normalisation (manual design)
Since there is a strong association of case control status with Source and flowcell information we are manually including them. Also, to be consistent with other brain banks we are including AgeAtDeath as a covariate.
```{r manual.adjusted.voom.normalisation, results='asis', echo=FALSE}
postAdjustCovars = c(postAdjustCovars, c('Source','FLOWCELL','AgeAtDeath')) %>% unique
randomCovariates = 'Donor_ID'

writeLines(paste('Using following covariates in the model:', 
                 paste(postAdjustCovars, collapse=', '),
                 'as fixed effects and',
                 paste(randomCovariates, collapse=', '),
                 'random effect'))

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

@@

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

plotDendroAndColors(sampleTrees, 
                    colors = col.mat, 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[clust == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[clust == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```

@@

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'ID') %>%
  dplyr::select(-PMI)
write.table(COVARIATES, file = 'MAYOcovariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MAYOcovariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

@@

# Design Matrix
DM1$design = DM1$design[,!(colnames(DM1$design) %in% c('PMI'))]
write.table(DM1$design, file='MAYOdesign.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MAYOdesign.tsv', 
                  name = 'Design Matrix',
                  parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ,
                      used = ALL_USED_IDs, 
                      activityName = activityName, 
                      executed = thisFile, 
                      activityDescription = activityDescription)
